Cognate beginnings to bilingual lexical acquisition

### 2023-07-03 manuscript/manuscript.tex
Add comments by DAV

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{tipa}
\usepackage{booktabs}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Cognate beginnings to bilingual lexical acquisition},
  pdfauthor={Gonzalo Garcia-Castro; Daniela S. Avila-Varela; Ignacio Castillejo; Nuria Sebastian-Galles},
  pdflang={en},
  pdfkeywords={cognate, word acquisition, vocabulary, bilingualism, item
response theory, bayesian},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\doublespacing
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Cognate beginnings to bilingual lexical
acquistion }
\title{Cognate beginnings to bilingual lexical acquisition}
\author{
\textbf{Gonzalo Garcia-Castro}~\orcidlink{0000-0002-8553-4209}\\Center
for Brain and Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\href{mailto:gonzalo.garciadecastro@upf.edu}{gonzalo.garciadecastro@upf.edu}\\\\\\
\textbf{Daniela S. Avila-Varela}~\orcidlink{0000-0002-3518-8117}\\Center
for Brain and Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\href{mailto:avila.varela.daniela@gmail.com}{avila.varela.daniela@gmail.com}\\\\\\
\textbf{Ignacio
Castillejo}~\orcidlink{0000-0001-7445-0416}\\Departamento de
Psicología\\Universidad Autónoma de Madrid\\Madrid,
28049\\\href{mailto:jignaciocastillejo@gmail.com}{jignaciocastillejo@gmail.com}\\\\\\
\textbf{Nuria Sebastian-Galles}~\orcidlink{0000-0001-6938-2498}\\Center
for Brain and Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\href{mailto:nuria.sebastian@upf.edu}{nuria.sebastian@upf.edu}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Bilingual infants show equivalent developmental trajectories of lexical
acquisition compared to their monolingual peers. This is remarkable,
given the increased complexity of their linguistic input. Recent studies
suggested that bilingual vocabulary growth is boosted by the number of
cognates shared by the pair of languages being learned, and that this
facilitation effect is driven by a stronger parallel activation of
cognates during linguistic exposure, compared to non-cognates. The
mechanisms behind this facilitation are still unclear. In this study, we
capitalise on accumulator models of language acquisition to propose an
account of bilingual lexical acquisition in which parallel activation
increases the rate at which children accumulate learning instances for
words in both languages, even in fully monolingual situations. Under
this hypothesis, we predicted a stronger cognate facilitation for words
to which children were exposed less frequently (low-exposure words), as
they are co-activated by their translation more often than high-exposure
words do. We developed an extensive online vocabulary checklist, the
Barcelona Vocabulary questionnaire (BVQ) to collect vocabulary data from
366 Catalan-Spanish bilingual toddlers aged 12 to 32 months. We then
used Bayesian explanatory item response theory to model the acquisition
trajectories of 302 Catalan/Spanish translation equivalents. We found an
interaction between exposure and cognateness, which pointed to
cognateness facilitating the aquisition of low-exposure words, but not
of mean exposure or high-exposure words. Overall, our findings suggest
that cognateness plays a key role in bilingual lexical acquisition, and
provides evidence for a frequency-mediated facilitation effect driven by
parallel activation.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
cognate \sep word
acquisition \sep vocabulary \sep bilingualism \sep item response
theory \sep 
bayesian

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, frame hidden, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi

\hypertarget{introductionauthor-note}{%
\section[Introduction]{\texorpdfstring{Introduction\footnote{The authors
  declare no conflicts of interest with regard to the funding source of
  this study. This research was supported by grants from the Spanish
  Ministerio de Ciencia, Innovación y Universidades
  (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government
  {[}ICREA (Catalan Institution for Research and Advanced Studies)
  Academia 2019 award{]}. Gonzalo Garcia-Castro was supported by a
  fellowship of the Spanish Ministerio de Ciencia, Innovación y
  Universidades (FPI 2019). We are grateful to Chiara Santolin, Ege E.
  Özer, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and the rest
  of the Speech Acquisition and Perception research group for their
  helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina
  Cuadrado for their technical support, and Cristina Dominguez and Katia
  Pistrin for their efforts in recruiting infants. We also thank all
  families and infants who participated in the experiments.}}{Introduction}}\label{introductionauthor-note}}

@@

Floccia et al. (\protect\hyperlink{ref-floccia2018introduction}{2018})
published an impressive study comparing vocabularies of children
learning several language pairs differing in their percentage of
cognates. The authors collected vocabulary data on word comprehension
and production from 372 24-month-old bilingual toddlers living in the
United Kingdom who were learning English and an additional language. The
additional language was one of 13 typologically diverse languages:
Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu,
Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The
authors calculated the average phonological overlap between the words in
each of these additional languages and their translation equivalents in
English, which was taken as a proxy of the degree of cognateness between
each pair of languages. Floccia and co-workers reported an increase in
vocabulary size in the additional language (i.e., not English)
associated with an increase in the average phonological similarity
between the translation equivalents of each language pair. For example,
English-Dutch bilinguals (languages with a high phonological overlap),
were able to produce more Dutch words than English-Mandarin bilinguals
(languages with a low phonological overlap) were able to produce in
Mandarin. Blom et al.
(\protect\hyperlink{ref-blom2020crosslanguage}{2020}), Bosma et al.
(\protect\hyperlink{ref-bosma2019longitudinal}{2019}), and Gampe et al.
(\protect\hyperlink{ref-gampe2021does}{2021}) reported similar results,
providing converging evidence of a facilitatory effect of language
distance on vocabulary size.

@@

Parallel activation of bilinguals' lexicons has been proposed as the
underlying mechanism for such facilitatory effect (e.g.,
\protect\hyperlink{ref-floccia2018introduction}{Floccia et al., 2018};
\protect\hyperlink{ref-mitchell2022cognates}{Mitchell et al., 2022}).
The parallel activation hypothesis stems from the language non-selective
account of lexical access, which suggests that bilinguals activate both
languages simultaneously during language processing, even in fully
monolingual contexts Evidence with adult bilinguals supporting the
language-non selective account of lexical access has been reported for
language comprehension and production, across the auditory and visual
(reading and signing) modalities
(\protect\hyperlink{ref-gimeno-martinez2021crosslinguistic}{Gimeno-Martínez
et al., 2021}; \protect\hyperlink{ref-vanhell2008sentence}{Hell \&
Groot, 2008}; \protect\hyperlink{ref-hoshino2008cognate}{Hoshino \&
Kroll, 2008}; \protect\hyperlink{ref-morford2011when}{Morford et al.,
2011}; \protect\hyperlink{ref-schwartz2007reading}{Schwartz et al.,
2007}; \protect\hyperlink{ref-spivey1999cross}{Spivey \& Marian, 1999};
see \protect\hyperlink{ref-kroll2017bilingual}{Kroll \& Ma, 2017} for
review). One of the clearest pieces of evidence of parallel activation
was provided by Costa et al.
(\protect\hyperlink{ref-costa2000cognate}{2000}). In this study, Spanish
monolinguals and Catalan-Spanish bilingual adults were asked to name
pictures of common objects in Spanish. In half of the trials, the object
labels were cognates in Spanish and Catalan (\emph{árbol}-\emph{arbre},
translations of \emph{tree}), whereas in the other half of the trial
labels were non-cognates (\emph{mesa}-\emph{taula}, translations of
\emph{table}), obviously, such distinction was only relevant for
bilinguals. Bilinguals named cognate pictures faster than non-cognate
pictures, even after adjusting for the lexical frequency of the items.
In contrast, Spanish monolinguals, who were unfamiliar with the Catalan
translations of the Spanish words they uttered, showed equivalent naming
times for the two types of stimuli. The authors interpreted the
difference between cognates and non-cognates in bilinguals as reflecting
the additional phonological activation that cognate words would receive
from their translation equivalents (due to language non-selective
activation of bilinguals' lexicons). These results showed that
bilinguals' Catalan phonology was activated during the production of
Spanish words, facilitating the naming of cognate pictures. Evidence of
parallel activation has been reported in bilingual toddlers and children
(\protect\hyperlink{ref-bosma2020cognate}{Bosma \& Nota, 2020};
\protect\hyperlink{ref-poarch2012crosslanguage}{Poarch \& Hell, 2012};
\protect\hyperlink{ref-poulin-dubois2013lexical}{Poulin-Dubois et al.,
2013}; \protect\hyperlink{ref-schroter2016orthographic}{Schröter \&
Schroeder, 2016}; \protect\hyperlink{ref-vonholzen2019impact}{Von Holzen
et al., 2019}; \protect\hyperlink{ref-vonholzen2012language}{Von Holzen
\& Mani, 2012}).

@@

We propose an account in which a learning instance for a word may also
represent a learning instance for its translation equivalent, to the
extent that such translation equivalent is co-activated\footnote{We use
  the term \emph{learning instance} in the fashion of accumulator models
  of language acquisition; as an exposure to a word-form that
  constitutes an opportunity for the child to accumulate information
  about the word. We do not make strong assumptions about whether a
  learning instance is a discrete or a continuous unit of accumulation
  of information. We rather consider that a learning instance of a word
  is so to the degree of the strength of activation of its phonological
  representation. This activation may be the result of the infant being
  exposed to the word-form, or the result of activation spreading
  through phonological or semantic links across lexicl representation,
  as in the case of parallel activation.}. The strength of this
co-activation is proportional to the phonological similarity between the
two translation equivalents; given that cognates share higher
phonological similarity than non-cognates, the former should be
co-activated more frequently than the latter. This should lead to a
faster accumulation of learning instances for cognates, compared to
non-cognates. Parallel activation would allow bilingual children to
accumulate learning instances for words in both languages even during
fully monolingual situations, but the impact of this mechanism would be
asymmetric across languages: words from the lower-exposure language
would receive additional activation through parallel activation more
often than words from the higher-exposure language. Therefore, the
acquisition of words from the lower-exposure language would benefit more
strongly from their cognate status than words from the higher-exposure
language.

Consider the example of the Catalan-Spanish cognate translation
equivalent /\textipa{"gat}-\textipa{"ga.to}/ {[}cat{]}, that are
phonologically very similar. When the child is exposed to
/\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in
parallel. Therefore, this exposure will count as a learning instance for
both co-activated forms. However, for the case of the non-cognate
translation equivalent /\textipa{"gos}/, the low phonological similarity
will result in a weak activation of /\textipa{"pe.ro}/ resulting in such
exposure counting as a learning instance for /\textipa{"gos}/ (which the
child was exposed to), but not for /\textipa{"pe.ro}/. While
/\textipa{"gat}-\textipa{"ga.to}/ will benefit from parallel activation,
/\textipa{"gos}-\textipa{"pe.ro}/ will not. If the child receives
linguistic input from one of the languages more often than from the
other, this effect might affect each form of the cognate translation
equivalent differently. For instance, if the child receives a larger
amount of Catalan input than Spanish input, they will encounter the
Catalan form /\textipa{"gat}/ more frequently than the Spanish form
/\textipa{"ga.to}/. Through parallel activation, /\textipa{"gat}/ will
activate /\textipa{"ga.to}/ more often than vice versa. Ultimately,
/\textipa{"ga.to}/ will benefit more strongly from its cognate status
than /\textipa{"gat}/, as it receives additional learning instances from
its translation equivalent more often than /\textipa{"gat}/.

To test these predictions, we collected vocabulary data on production
and comprehension from a large sample of bilingual Catalan-Spanish
children using an online vocabulary checklist designed for the present
study. We adopted a Bayesian explanatory item response theory approach
(IRT, see \protect\hyperlink{ref-kachergis2022standard}{Kachergis et
al., 2022}, for a similar approach) to model the probability of
acquisition of 604 Catalan and Spanish nouns included in the vocabulary
checklist. Words were considered as acquired if caregivers reported such
word to be understood (comprehension) or understood and said
(production) by their child. We estimated the impact of several
predictors of interest on the probability of acquisition, including
participants' age and rate of exposure to the word-form, and the cognate
status of the word-form. As described in the methods section, rate of
exposure was a composite measure taking into account participant'
language exposure and word's lexical frequency. We predicted an
interaction between cognate status and word exposure rate in which the
probability of comprehension is higher for low-exposure cognate words,
but not for high-exposure words cognate words.

@@

\hypertarget{tbl-items}{}
\begin{longtable}{l|rrrrc}
\caption{\label{tbl-items}Summary of the items included in the final analyses. }\tabularnewline

\toprule
\multicolumn{1}{l}{} & List A & List B & List C & List D & Examples \\ 
\midrule
\multicolumn{6}{l}{Semantic category} \\ 
\midrule
Household items & 31 & 26 & 30 & 25 & clock, video \\ 
Food and drink & 29 & 26 & 23 & 27 & sausage, yogurt \\ 
Animals & 26 & 23 & 19 & 25 & panther, tiger \\ 
Outside & 14 & 13 & 13 & 15 & farm, stone \\ 
Body parts & 14 & 12 & 11 & 11 & face, finger \\ 
Toys & 11 & 11 & 12 & 13 & piano, racket \\ 
Clothes & 12 & 12 & 10 & 10 & zipper, sandal \\ 
Vehicles & 9 & 10 & 11 & 10 & helicopter, tractor \\ 
Colours & 6 & 6 & 6 & 6 & red, green \\ 
People & 7 & 4 & 6 & 6 & police, babysitter \\ 
Furniture and rooms & 4 & 4 & 4 & 4 & corridor, terrace \\ 
Time & 2 & 2 & 2 & 2 & day, night \\ 
Adventures & 1 & 1 & 1 & 1 & witch \\ 
Parts of things & 1 & 1 & 1 & 1 & wheel \\ 
\midrule 
Total & 167 & 151 & 149 & 156 & — \\ 
\bottomrule
\end{longtable}

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{tipa}
\usepackage{booktabs}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Cognate beginnings to bilingual lexical acquisition},
  pdfauthor={Gonzalo Garcia-Castro; Daniela S. Avila-Varela; Ignacio Castillejo; Nuria Sebastian-Galles},
  pdflang={en},
  pdfkeywords={cognate, word acquisition, vocabulary, bilingualism, item
response theory, bayesian},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Cognate beginnings to bilingual lexical
acquistion }
\title{Cognate beginnings to bilingual lexical acquisition\thanks{The
authors declare no conflicts of interest with regard to the funding
source of this study. GGC was supported by the This research was
supported by a FPI 2020 research contract (PRE2019-088165) in a project
funded by the Spanish Ministry of Science, Innovation, and Universities
(PGC2018-101831-B-I00). DAV was supported by the European Union's
Horizon 2023 research and innovation program under Marie
Skłodowska--Curie Grant (765556) and a postdoctoral fellowship from the
Foundation for Science and Technology of Portugal (UIDB/00214/2020). NSG
was supported by Academia 2019 award from the Catalan Institution for
Research and Advanced Studies (ICREA). We are grateful to Chiara
Santolin, Ege E. Özer, Alicia Franco-Martínez, Cristina Rodríguez-Prada,
and the rest of the Speech Acquisition and Perception research group for
their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and
Cristina Cuadrado for their technical support, and Cristina Dominguez
and Katia Pistrin for their efforts in recruiting infants. We also thank
all families and infants who participated in the experiments.}}
\author{
\textbf{Gonzalo Garcia-Castro}~\orcidlink{0000-0002-8553-4209}\\Center
for Brain and Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\href{mailto:gonzalo.garciadecastro@upf.edu}{gonzalo.garciadecastro@upf.edu}\And
\textbf{Daniela S. Avila-Varela}~\orcidlink{0000-0002-3518-8117}\\Center
for Brain and Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\href{mailto:avila.varela.daniela@gmail.com}{avila.varela.daniela@gmail.com}\And
\textbf{Ignacio
Castillejo}~\orcidlink{0000-0001-7445-0416}\\Departamento de
Psicología\\Universidad Autónoma de Madrid\\Madrid,
28049\\\href{mailto:jignaciocastillejo@gmail.com}{jignaciocastillejo@gmail.com}\And
\textbf{Nuria Sebastian-Galles}~\orcidlink{0000-0001-6938-2498}\\Center
for Brain and Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\href{mailto:nuria.sebastian@upf.edu}{nuria.sebastian@upf.edu}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Bilingual infants show equivalent developmental trajectories of lexical
acquisition compared to their monolingual peers. This is remarkable,
given the increased complexity of their linguistic input. Recent studies
suggested that bilingual vocabulary growth is boosted by the number of
cognates shared by the pair of languages being learned, and that this
facilitation effect is driven by a stronger parallel activation of
cognates during linguistic exposure, compared to non-cognates. The
mechanisms behind this facilitation are still unclear. In this study, we
capitalise on accumulator models of language acquisition to propose an
account of bilingual lexical acquisition in which parallel activation
increases the rate at which children accumulate learning instances for
words in both languages, even in fully monolingual situations. Under
this hypothesis, we predicted a stronger cognate facilitation for words
to which children were exposed less frequently (low-exposure words), as
they are co-activated by their translation more often than high-exposure
words do. We developed an extensive online vocabulary checklist, the
Barcelona Vocabulary questionnaire (BVQ) to collect vocabulary data from
366 Catalan-Spanish bilingual toddlers aged 12 to 32 months. We then
used Bayesian explanatory item response theory to model the acquisition
trajectories of 302 Catalan/Spanish translation equivalents. We found an
interaction between exposure and cognateness, which pointed to
cognateness facilitating the aquisition of low-exposure words, but not
of mean exposure or high-exposure words. Overall, our findings suggest
that cognateness plays a key role in bilingual lexical acquisition, and
provides evidence for a frequency-mediated facilitation effect driven by
parallel activation.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
cognate \sep word
acquisition \sep vocabulary \sep bilingualism \sep item response
theory \sep 
bayesian

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, sharp corners, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

Floccia et al. (\protect\hyperlink{ref-floccia2018introduction}{2018})
published an impressive study comparing vocabularies of children
learning several language pairs differing in their percentage of
cognates. The authors collected vocabulary data on word comprehension
and production from 372 24-month-old bilingual toddlers living in the
United Kingdom who were learning English and an additional language. The
additional language was one of 13 typologically diverse languages:
Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu,
Italian, Mandarin Chinese, Polish, Portuguese, Spanish, and Welsh. The
authors calculated the average phonological overlap between the words in
each of these additional languages and their translation equivalents in
English, which was taken as a proxy of the degree of cognateness between
each pair of languages. Floccia and co-workers reported an increase in
vocabulary size in the additional language (i.e., not English)
associated with an increase in the average phonological similarity
between the translation equivalents of each language pair. For example,
English-Dutch bilinguals (languages with a high phonological overlap),
were able to produce more Dutch words than English-Mandarin bilinguals
(languages with a low phonological overlap) were able to produce in
Mandarin. Blom et al.
(\protect\hyperlink{ref-blom2020crosslanguage}{2020}), Bosma et al.
(\protect\hyperlink{ref-bosma2019longitudinal}{2019}), and Gampe et al.
(\protect\hyperlink{ref-gampe2021does}{2021}) reported similar results,
providing converging evidence of a facilitatory effect of a lower
language distance (i.e., higher degre of cognateness) on vocabulary
size.

@@

Parallel activation of bilinguals' lexicons has been proposed as the
underlying mechanism for such facilitatory effect (e.g.,
\protect\hyperlink{ref-floccia2018introduction}{Floccia et al., 2018};
\protect\hyperlink{ref-mitchell2022cognates}{Mitchell et al., 2022}).
The parallel activation hypothesis stems from the language non-selective
account of lexical access, which suggests that bilinguals activate both
languages simultaneously during language processing, even in fully
monolingual contexts. Evidence with adult bilinguals supporting the
language-non selective account of lexical access has been reported for
language comprehension and production, across the auditory and visual
(reading and signing) modalities
(\protect\hyperlink{ref-gimeno-martinez2021crosslinguistic}{Gimeno-Martínez
et al., 2021}; \protect\hyperlink{ref-vanhell2008sentence}{Hell \&
Groot, 2008}; \protect\hyperlink{ref-hoshino2008cognate}{Hoshino \&
Kroll, 2008}; \protect\hyperlink{ref-morford2011when}{Morford et al.,
2011}; \protect\hyperlink{ref-schwartz2007reading}{Schwartz et al.,
2007}; \protect\hyperlink{ref-spivey1999cross}{Spivey \& Marian, 1999};
see \protect\hyperlink{ref-kroll2017bilingual}{Kroll \& Ma, 2017} for
review). One of the clearest pieces of evidence of parallel activation
was provided by Costa et al.
(\protect\hyperlink{ref-costa2000cognate}{2000}). In this study, Spanish
monolinguals and Catalan-Spanish bilingual adults were asked to name
pictures of common objects in Spanish. In half of the trials, the object
labels were cognates in Spanish and Catalan (\emph{árbol}-\emph{arbre},
translations of \emph{tree}), whereas in the other half of the trial
labels were non-cognates (\emph{mesa}-\emph{taula}, translations of
\emph{table}), obviously, such distinction was only relevant for
bilinguals. Bilinguals named cognate pictures faster than non-cognate
pictures, even after adjusting for the lexical frequency of the items.
In contrast, Spanish monolinguals, who were unfamiliar with the Catalan
translations of the Spanish words they uttered, showed equivalent naming
times for the two types of stimuli. The authors interpreted the
difference between cognates and non-cognates in bilinguals as reflecting
the additional phonological activation that cognate words would receive
from their translation equivalents (due to language non-selective
activation of bilinguals' lexicons). These results showed that
bilinguals' Catalan phonology was activated during the production of
Spanish words, facilitating the naming of cognate pictures. Evidence of
parallel activation has been reported in bilingual toddlers and children
(\protect\hyperlink{ref-bosma2020cognate}{Bosma \& Nota, 2020};
\protect\hyperlink{ref-poarch2012crosslanguage}{Poarch \& Hell, 2012};
\protect\hyperlink{ref-poulin-dubois2013lexical}{Poulin-Dubois et al.,
2013}; \protect\hyperlink{ref-schroter2016orthographic}{Schröter \&
Schroeder, 2016}; \protect\hyperlink{ref-vonholzen2019impact}{Von Holzen
et al., 2019}; \protect\hyperlink{ref-vonholzen2012language}{Von Holzen
\& Mani, 2012}).

@@

We propose an account in which a learning instance for a word may also
represent a learning instance for its translation equivalent, to the
extent that such translation equivalent is co-activated\footnote{We use
  the term \emph{learning instance} in the fashion of accumulator models
  of language acquisition; as an exposure to a word-form that
  constitutes an opportunity for the child to accumulate information
  about the word. We do not make strong assumptions about whether a
  learning instance is a discrete or a continuous unit of accumulation
  of information. We rather consider that a learning instance of a word
  is so to the degree of the strength of activation of its phonological
  representation. This activation may result from the infant being
  exposed to the word-form, or the result of activation spreading
  through phonological or semantic links across lexical representation,
  as in the case of parallel activation.}. The strength of this
co-activation is proportional to the phonological similarity between the
two translation equivalents; given that cognates share higher
phonological similarity than non-cognates, the former should be
co-activated more frequently than the latter. This should lead to a
faster accumulation of learning instances for cognates, compared to
non-cognates. Parallel activation would allow bilingual children to
accumulate learning instances for words in both languages even during
fully monolingual situations, but the impact of this mechanism would be
asymmetric across languages: words from the lower-exposure language
(i.e., the language to which the infants are exposed to more often)
would receive additional activation through parallel activation more
frequently than words from the higher-exposure language. Therefore, the
acquisition of words from the lower-exposure language would benefit more
strongly from their cognate status than words from the higher-exposure
language.

Consider the example of the Catalan-Spanish cognate translation
equivalent /\textipa{"gat}-\textipa{"ga.to}/ {[}cat{]}, that are
phonologically very similar. When the child is exposed to
/\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in
parallel. Therefore, this exposure will count as a learning instance for
both co-activated forms. The case of the non-cognate translation
equivalent /\textipa{"gos-"pe.ro}/ would be different. Given the low
phonological similarity between both word-forms, an exposure to
\textipa{"gos} will result in a weak activation of /\textipa{"pe.ro}/
leading to such exposure counting as a learning instance for
/\textipa{"gos}/ (which the child was exposed to), but not for
/\textipa{"pe.ro}/. While /\textipa{"gat}-\textipa{"ga.to}/ will benefit
from parallel activation, /\textipa{"gos}-\textipa{"pe.ro}/ will not. If
the child receives linguistic input from one of the languages more often
than from the other, this effect might affect each form of the cognate
translation equivalent differently. For instance, if the child receives
a larger amount of Catalan input than Spanish input, they will encounter
the Catalan form /\textipa{"gat}/ more frequently than the Spanish form
/\textipa{"ga.to}/. Through parallel activation, /\textipa{"gat}/ will
activate /\textipa{"ga.to}/ more often than vice versa. Ultimately,
/\textipa{"ga.to}/ will benefit more strongly from its cognate status
than /\textipa{"gat}/, as it receives additional learning instances from
its translation equivalent more often than /\textipa{"gat}/.

To test these predictions, we collected vocabulary data on production
and comprehension from a large sample of bilingual Catalan-Spanish
children using the Barcelona Vocabulary Questionnaire (BVQ)
(\protect\hyperlink{ref-garcia-castro2023bvq}{Garcia-Castro et al.,
2023}), a dual language online vocabulary checklist designed for the
present study. We adopted a Bayesian explanatory item response theory
approach (IRT, see
\protect\hyperlink{ref-kachergis2022standard}{Kachergis et al., 2022},
for a similar approach) to model the probability of acquisition of 604
Catalan and Spanish nouns included in the vocabulary checklist. Words
were considered as acquired if caregivers reported such word to be
understood (comprehension) or understood and said (production) by their
child. We estimated the impact of several predictors of interest on the
probability of acquisition, including participants' age and rate of
exposure to the word-form, and the cognate status of the word-form. As
described in the methods section, rate of exposure was a composite
measure taking into account participant' language exposure and word's
lexical frequency. We predicted an interaction between cognate status
and word exposure rate in which the probability of comprehension is
higher for low-exposure cognate words, but not for high-exposure cognate
words.

@@

\hypertarget{tbl-items}{}
\begin{table}
\caption{\label{tbl-items}Summary of the items included in the final analyses. }\tabularnewline

\centering
\begin{tabular}{lccccl}
\toprule
Semantic category & List A & List B & List C & List D & Examples\\
\midrule
Household items & 31 & 26 & 30 & 25 & clock, video\\
Food and drink & 29 & 26 & 23 & 27 & sausage, yogurt\\
Animals & 26 & 23 & 19 & 25 & panther, tiger\\
Outside & 14 & 13 & 13 & 15 & farm , stone\\
Body parts & 14 & 12 & 11 & 11 & face  , finger\\
\addlinespace
Toys & 11 & 11 & 12 & 13 & piano , racket\\
Clothes & 12 & 12 & 10 & 10 & zipper, sandal\\
Vehicles & 9 & 10 & 11 & 10 & helicopter, tractor\\
Colours & 6 & 6 & 6 & 6 & red  , green\\
People & 7 & 4 & 6 & 6 & police    , babysitter\\
\addlinespace
Furniture and rooms & 4 & 4 & 4 & 4 & corridor, terrace\\
Time & 2 & 2 & 2 & 2 & day  , night\\
Adventures & 1 & 1 & 1 & 1 & witch\\
Parts of things & 1 & 1 & 1 & 1 & wheel\\
\midrule
N & 167 & 151 & 149 & 156 & \\
\bottomrule
\end{tabular}
\end{table}
